Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-04-03 Manuscript/manuscript.Rmd
results edits

## Random under-sampling, distance-based predictors, and 10 kb resolutions of Hi-C data provided the optimal data level characteristics for TAD boundary prediction

Expectedly, when using data with class imbalance present, that is no re-sampling, the models exhibited low MCCs, with very little variability among different resolutions (Figure 3). Similarly poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS narrowly out performed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction. <!--These results suggest that a combination of both random over- and under-sampling (SMOTE) can improve TAD boundary prediction.-->Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type feature space yielded substantially higher MCCs when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with comparing class balancing techniques, this improvement was less evident at lower resolutions. <!--This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations.-->

Interestingly, for the optimal re-sampling and predictor type, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the sparsity in the interaction frequency matrices at finer resolutions. This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Furthermore, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). In summary, our computational framework showed that using random under-sampling, distance-type predictors, and Hi-C data at 10kb resolution yields an optimally performing cell line-specific random forest classification model for TAD boundary prediction.

![**Models built using distance-type predictors combined with random under-sampling at 10 kb yielded optimal predictive performance for GM12878.** Line plots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Performances are compared across resolution, within each predictor-type: (A) OC, (B) OP, and (C) Distance, and between each re-sampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/performances.png)

>>>>

## 5 kb resolution, random under-sampling, and distance-based predictors provided optimal performance for TAD boundary prediction

<!--
Expectedly, when using data with class imbalance present, that is no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS narrowly out performed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction. Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type feature space yielded substantially higher MCCs when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with comparing class balancing techniques, this improvement was less evident at lower resolutions.

Interestingly, for the optimal re-sampling and predictor type, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the sparsity in the interaction frequency matrices at finer resolutions. This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Furthermore, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). In summary, our computational framework showed that using random under-sampling, distance-type predictors, and Hi-C data at 10kb resolution yields an optimally performing cell line-specific random forest classification model for TAD boundary prediction.
-->

![**Models built using distance-type predictors combined with random under-sampling at 5 kb yielded optimal predictive performance for GM12878.** Line plots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Performances are compared across resolution, within each predictor-type: (A) OC, (B) OP, and (C) Distance, and between each re-sampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Performances were evaluated using balanced accuracy.](Figures/performances.png)
